### 2017-12-06 manuscript.rmd
small ms changes, new knit

Siphonophore colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and there is a single gain of monoecy within the Codonophora; however the lack of resolution of deep relationships within Codonophora made it impossible to rule out alternative evolutionary scenarios [@Dunn:2005dy]. The current better-resolved tree (Fig. 5A), indicates that monoecy in siphonophores evolved twice from a dioecious ancestor, in the branch leading to Calycophorae and in the branch leading to Agalmatids *sensu lato*. There is a small probability for an alternative scenario featuring a single gain of monoecy before the split of the Codonophora, with a subsequent shift to dioecy in the *Marrus-Erenna* clade.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969;@cartwright2010character]. For example, the siphonophore genus *Forskalia* has 6 basic zooid types (pneumatophore, nectophore, gastrozooid, palpon, bract, and gonophore), and a total of 10 counting subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Diphyomorphs have more than 1 type of propulsive nectophore, while cystonects have none. Here we reconstruct the evolutionary origins of the different zooid types on the present transcriptome tree (Fig. 4).

Nectophores are retained modified medusae that codonophorans use for coordinated colony-level swimming. The nectosome is the region of the colony that develops from the nectosomal growth zone (Fig. 2B). Unlike the siphosomal growth zone, the nectosomal growth zone does not bud gastrozooids, but nectophores (and in the case of *Apolemia*, also palpons). Nectophores are exclusively found on the nectosome -with the exception of  *Physalia physalis* (which has no nectosome, and grows small nectophores near the gonodendra [@Totton1960]). It is possible that the common ancestor of siphonophores had a nectosome, which was lost on the branch leading to cystonects. We cannot exclude with certainty the alternative hypothesis of a nectosome-less ancestor followed by a gain of the nectosome in the branch leading to Codonophora. It is suggested that the nectosome arose as a duplication of the siphosome, followed by functional specialization in propelling the colony [@dunn2006evolution]. The nectosome has been lost within Codonophora in the genus *Athorybia*.

@@

Bracts are highly reduced zooids unique to siphonophores, but they are only present in Codonophora. As with the nectosome, we have ambiguity when determining whether the common ancestor of siphonophores had bracts or not. The common ancestor of Codonophora had only one bract subtype, which was lost in Hippopodidae, *Physophora hydrostatica* (however, they are present in its sister species, *P. gilmeri* included in the present phylogeny), and in *Gymnopraia lapislazula* (not included in present phylogeny). Bracts are functional for protection of the delicate zooids and to help maintain neutral buoyancy [@jacobs1937beobachtungen]. Some calycophorans are able to actively exclude sulphate ions in their bracts to adjust their buoyancy along the colony [@bidigare1980role].

Palpons are modified mouthless gastrozooids used for digestion and circulation of the gastrovascular fluid [@mackie1987siphonophore]. They were present in the common ancestor of siphonophores (Fig. 5B), retained in most species, but lost three times independently in the branches leading to *Pyrostephidae* (represented here by the genera *Bargmannia* and *Rudjakovia*), in calycophorans, and in *Marrus claudanielis*. These taxa might have found other avenues to effectively circulate nutrients across the colony.

@@

The pneumatophore is a gas filled float located at the anterior end of the colony, which helps the colony float and maintain its orientation in the water column [@church2015histology;@mackie1974locomotion]. Recent evidence of neural arrangement in the pneumatophore of *Nanomia bijuga*  suggests it could also gather information on relative pressure changes (and thus depth changes), helping regulate geotaxis [@church2015histology]. The ancestral siphonophore had a pneumatophore (Fig. 2B), since both cystonects and most codonophorans possess one (Fig. 4). It is not considered a zooid, as it is not formed by budding but by invagination at the aboral end of the planula during early development [@garstang1946morphology; @leloup1935siphonophores; @carre1969etude].  The pneumatophore was lost in Calycophorae and never gained again in that clade. Calycophorans rely on the ionic balance of their gelatinous nectophores and bracts to retain posture and neutral buoyancy [@mackie1974locomotion].

@@

Siphonophores are abundant predators in the pelagic realm, ranging from the surface (*Physalia physalis*) to bathypelagic depths (Fig. 4, S6) [@mapstone2014global]. However, the depth distribution of siphonophore populations is not always static. Many species are strong vertical migrators, and some species such as *Chuniphyes multidentata* have exhibited synchronous diel migration patterns [@pugh1984diel]. The phylogeny indicates that the siphonophore common ancestor was probably planktonic, as most extant siphonophore taxa are. Some siphonophore species (such as *Bargmannia lata*) have been observed at abyssopelagic depths near 4000m. Using the present phylogeny, we reconstructed the median depth changes along the phylogeny under a Brownian Motion model (Fig. S6). This model indicates a mesopelagic common ancestor, with several independent transition events to epipelagic (light blue) and bathypelagic (dark blue) waters. This reconstruction only included species recorded using an ROV, thus it excludes many other independent colonizations of the epipelagic habitat. The ROV observations are reliable below 200m, and no quantative measurements were made on SCUBA dives. Species such as *Hippopodius hippopus*, *Athorybia rosacea*, *Diphyes dispar*, and *Chelophyes appendiculata* are often encountered blue water diving less than 20m from the surface (Fig. 4). There was only a single transition to benthic lifestyle on the branch of *Rhodaliidae*, and a single transition to a pleustonic lifestyle on the branch of *Physalia physalis*.

@@

This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award). Sequencing at the Brown Genomics Core facility was supported in part by NIH P30RR031153 and NSF EPSCoR EPS-1004057. Data transfer was supported by NSF RII-C2 EPS-1005789. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. SOWHAT analyses were carried out on the Odyssey cluster supported by the FAS Division of Science, Research Computing Group at Harvard University -- we thank Cassandra Extavour for use of Harvard cluster. We also thank the MBARI crews and ROV pilots for collection of the specimens.

>>>>

Siphonophore colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and there is a single gain of monoecy within the Codonophora; however the lack of resolution of deep relationships within Codonophora made it impossible to rule out alternative evolutionary scenarios [@Dunn:2005dy]. The current better-resolved tree (Fig. 5A), indicates that monoecy in siphonophores evolved twice from a dioecious ancestor, in the branch leading to Calycophorae and in the branch leading to [**placeholder name** Agalmatids *sensu lato*]. There is a small probability for an alternative scenario featuring a single gain of monoecy before the split of the Codonophora, with a subsequent shift to dioecy in the *Marrus-Erenna* clade.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969;@cartwright2010character]. For example, the genus *Forskalia* has 6 basic zooid types (pneumatophore, nectophore, gastrozooid, palpon, bract, and gonophore), and a total of 10 counting subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Diphyomorphs have more than 1 type of propulsive nectophore, while cystonects have none. Here we reconstruct the evolutionary origins of the different zooid types on the present transcriptome tree (Fig. 4).

Nectophores are retained modified medusae that codonophorans use for coordinated colony-level swimming. The nectosome is the region of the colony that develops from the nectosomal growth zone (Fig. 2B). Unlike the siphosomal growth zone, the nectosomal growth zone does not bud gastrozooids, but nectophores (and in the case of *Apolemia*, also palpons). It is possible that the common ancestor of siphonophores had a nectosome, which was lost on the branch leading to cystonects. We cannot exclude with certainty the alternative hypothesis of a nectosome-less ancestor followed by a gain of the nectosome in the branch leading to Codonophora. It is suggested that the nectosome arose as a duplication of the siphosome, followed by functional specialization in propelling the colony [@dunn2006evolution]. The nectosome has been lost within Codonophora in the genus *Athorybia*.

@@

Bracts are highly reduced zooids unique to siphonophores, but they are only present in Codonophora. Bracts are functional for protection of the delicate zooids and to help maintain neutral buoyancy [@jacobs1937beobachtungen]. Some calycophorans are able to actively exclude sulphate ions in their bracts to adjust their buoyancy along the colony [@bidigare1980role]. As with the nectosome, we cannot be sure whether the common ancestor of siphonophores had bracts or not. The common ancestor of Codonophora had only one bract subtype, which was lost in Hippopodidae, *Physophora hydrostatica*, the Clausophyidae, and in *Gymnopraia lapislazula* (none of the latter three species and clades are included in the phylogeny). Phylogenetic reconstructions are sensitive to taxon sampling, and these patterns of loss are not captured by the present sampling [**We don't have the reconstruction in supp info - do we want to add it?**].

Palpons are modified reduced gastrozooids used for digestion and circulation of the gastrovascular fluid [@mackie1987siphonophore]. We do not disinguish here between gonopalpons (palpons associated with gonodendra, without a tentacle, as in the cystonects) and palpons borne on the stem (typically with a reduced tentacle or palpacle) [@totton1965synopsis]. They were present in the common ancestor of siphonophores (Fig. 5B), retained in most species, but lost three times independently in the branches leading to *Pyrostephidae* (represented here by the genera *Bargmannia* and *Rudjakovia*), in calycophorans (with the exception of Stephanophyes [@totton1965synopsis]), and in *Marrus claudanielis*. These taxa might have found other avenues to effectively circulate nutrients across the colony.

@@

The pneumatophore is a gas filled float located at the anterior end of the colony, which helps the colony maintain its orientation in the water column, and plays a role in flotation in the case of the cystonects [@church2015histology;@mackie1974locomotion; @totton1965synopsis]. Recent evidence of neural arrangement in the pneumatophore of *Nanomia bijuga*  suggests it could also gather information on relative pressure changes (and thus depth changes), helping regulate geotaxis [@church2015histology]. The ancestral siphonophore had a pneumatophore (Fig. 2B), since both cystonects and most codonophorans possess one (Fig. 4). It is not considered a zooid, as it is not formed by budding but by invagination at the aboral end of the planula during early development [@garstang1946morphology; @leloup1935siphonophores; @carre1969etude].  The pneumatophore was lost in Calycophorae and never gained again in that clade. Calycophorans rely on the ionic balance of their gelatinous nectophores and bracts to retain posture and neutral buoyancy [@mackie1974locomotion].

@@

Siphonophores are abundant predators in the pelagic realm, ranging from the surface (*Physalia physalis*) to bathypelagic depths (Fig. 4, S6) [@mapstone2014global; @mackie1987siphonophore]. However, the depth distribution of siphonophore populations is not always static. Many species are strong vertical migrators, and some species such as *Chuniphyes multidentata* have exhibited synchronous diel migration patterns [@pugh1984diel]. The phylogeny indicates that the siphonophore common ancestor was probably planktonic, as most extant siphonophore taxa are. Some siphonophore species (such as *Bargmannia lata*) have been observed at abyssopelagic depths near 4000m. Using the present phylogeny, we reconstructed the median depth changes along the phylogeny under a Brownian Motion model (Fig. S6). This model indicates a mesopelagic common ancestor, with several independent transition events to epipelagic (light blue) and bathypelagic (dark blue) waters. This reconstruction only included species recorded using an ROV, thus it excludes many other independent colonizations of the epipelagic habitat. The ROV observations are reliable below 200m, and no quantative measurements were made on SCUBA dives. Species such as *Hippopodius hippopus*, *Athorybia rosacea*, *Diphyes dispar*, and *Chelophyes appendiculata* are often encountered blue water diving less than 20m from the surface (Fig. 4). There was only a single transition to benthic lifestyle on the branch of *Rhodaliidae*, and a single transition to a pleustonic lifestyle on the branch of *Physalia physalis*.

@@

This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award). Sequencing at the Brown Genomics Core facility was supported in part by NIH P30RR031153 and NSF EPSCoR EPS-1004057. Data transfer was supported by NSF RII-C2 EPS-1005789. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. SOWHAT analyses were carried out on the Odyssey cluster supported by the FAS Division of Science, Research Computing Group at Harvard University -- we thank Cassandra Extavour for use of the Harvard cluster. We also thank the MBARI crews and ROV pilots for collection of the specimens.
